---
title: "DATA612 - Project 1"
author: "Leland Randles"
date: "6/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Project 1
  
This system recommends clothing items based on ratings provided on other clothing items. The below user-item matrix, shows ratings for clothing items from 10 different users. Out of 50 user / clothes item combinations, 36 have been rated and 14 have not been rated.  

First the necessary packages were loaded:  
  
```{r ip}
library(caTools, quietly = TRUE)
library(dplyr, quietly = TRUE)
library(tidyverse, quietly = TRUE)
library(ModelMetrics, quietly = TRUE)
library(sqldf, quietly = TRUE)
```
  
<br>

#### Load and Prepare Data  
  
Then the ratings data was loaded and prepared:  
  
```{r data_prep}
clothes_ratings_raw <- read.csv("https://raw.githubusercontent.com/Randles-CUNY/DATA612/master/Proj_1/clothes_ratings.csv")
clothes_ratings_raw[clothes_ratings_raw=="?"] <- NA
clothes_ratings_raw <- data.frame(lapply(clothes_ratings_raw, as.character), stringsAsFactors=FALSE)
clothes_ratings_raw
```  
  
<br>
  
#### Split Data into Training and Test Datasets  
  
The `clothes_ratings_raw` dataset was then split into training and test datasets:  
  
```{r train_test}
# Lengthen data
clothes_ratings <- clothes_ratings_raw %>% gather(key = Item, value = Rating, -User)
clothes_ratings$Rating <- as.numeric(clothes_ratings$Rating)
set.seed(250)
train_test <- sample.split(clothes_ratings$Rating, SplitRatio = 0.75)
train <- clothes_ratings
train$Rating[!train_test] <- NA
test <- clothes_ratings
test$Rating[train_test] <- NA
```  
  
<br>
  
#### Compute Raw Average and RMSE for Both Datasets  
  
Having split the clothes ratings data into `train` and `test` datasets, the next steps were to compute the raw average for the `train` dataset and the RMSE for both the `train` and `test` data:  
  
```{r avg_rmse}
# Lengthen data
train_raw_avg <- mean(train$Rating, na.rm = TRUE)
train_raw_avg
train_raw_avg_vect_trn <- rep(train_raw_avg, times = length(which(!is.na(train$Rating))))
train_raw_avg_vect_tst <- rep(train_raw_avg, times = length(which(!is.na(test$Rating))))
train_rmse <- rmse(train$Rating[which(!is.na(train$Rating))], train_raw_avg_vect_trn)
train_rmse
test_rmse <- rmse(test$Rating[which(!is.na(test$Rating))], train_raw_avg_vect_tst)
test_rmse
```  
  
<br> 
  
#### Calculate Bias Values, Baseline Predictors and Associagted RMSEs  
  
Next, the training set was used to calculate bias values for each user and each clothing item, then baseline predictors were added for each user / clothing item combination. Finally, RMSE was calculated for the baseline predictors for both:  
  
```{r bias}
user_bias <- sqldf("select User, (sum(Rating)/count(Rating)) - 6 as user_bias from train where Rating <> 'NA' group by User")
clothes_item_bias <- sqldf("select Item, (sum(Rating)/count(Rating)) - 6 as clothes_item_bias from train where Rating <> 'NA' group by Item")
train <- merge(train, user_bias, by = "User")
train <- merge(train, clothes_item_bias, by = "Item")
train$baseline_pred <- 6 + train$user_bias + train$clothes_item_bias
train_baseline_pred_rmse <- rmse(train$baseline_pred[which(!is.na(train$Rating))], train_raw_avg_vect_trn)
train_baseline_pred_rmse
test <- merge(test, user_bias, by = "User")
test <- merge(test, clothes_item_bias, by = "Item")
test$baseline_pred <- 6 + test$user_bias + test$clothes_item_bias
test_baseline_pred_rmse <- rmse(test$baseline_pred[which(!is.na(test$Rating))], train_raw_avg_vect_tst)
test_baseline_pred_rmse
```  
  
<br>  
  
As you can see, the RMSE for the baseline predictors for both the `train` and `test` datasets dropped considerably compared to the RMSE which was calculated from raw averages.  